Type-Safe Modular Parsing

### 2017-01-12 paper/sections/RelatedWork.tex
Finished revision of Related Work

\paragraph{Parser Combinators} Since Burge~\cite{burge1975} firstly
introduced parser generators, and after Wadler popularized them 
in functional programming~\cite{Wadler1985}, parser combinators have been
more and more popularly developed and used in the research area of
parsing. Many parsing libraries produce recursive descent
parsers by introducing functional monadic
parser combinators~\cite{nott237}. Parsec~\cite{Leijen2001} is 
perhaps the most popular parser combinator library in this line. 
It is widely used in Haskell (with various ``clones'' in other languages)
for context-sensitive grammars with infinite lookahead. Nevertheless,
Parsec users suffer from manual left-recursion elimination, 
high cost for backtracking and longest match composition issues, 
as we discussed in Section~\ref{}. Those limitation make Parsec 
inadequate for modular parsing.

@@

\paragraph{Extensibility} To the best of our knowledge, none of the existing parsing techniques
address the problem of producing extensible ASTs. Certainly, there has been a lot of work on extensible
data types, in both functional and object-oriented programming.
For instance, the famous DTC paper~\cite{swierstra2008} represents
modular ASTs using co-products of every two functors, which forms a
binary tree structure. Later
Oliveira et al. presented MRM~\cite{Oliveira2015} in Haskell, which is
similar to DTC but adopts a list of functors in type-level lists
instead. Other related work
includes~\cite{Bahr2014}, where data types are flexible with
composition as well as decomposition.

In OO languages, however, things get much better with the mechanism of
inheritance. Furthermore, Object Algebras~\cite{Oliveira2012} were
proposed as a design pattern for the extensibility of both data
variants and operations in a modular type-safe way. The code is
concise and expressive using Object Algebras, building a foundation
for our new library with various dimensions of extensibility.
Moreover, in~\cite{Oliveira2012} the authors have discussed the
composition of algebras. In our parsing approach, a parser consumes an
algebra, which is delegated to return the results, during its process
of parsing. Having a set of algebras, it requires multiple parsing
with several times of invocation, which leads to redundant work.
Instead, algebras are supposed to be composed into one before the
invocation of the parser. Bahr et al. lead a similar discussion
in~\cite{Bahr2011}, where queries (or \textit{catamorphisms}) and
transformations (or \textit{homomorphisms}) are composable. They have
also mentioned the dual process of folds, namely
\textit{anamorphisms}. It is potentially related to our work, as
parsing is a representative kind of unfolds, whereas they only
discussed the composition of a cv-coalgebra and a term homomorphism,
which differs from modular parsing.

Another interesting observation, as we have mentioned
in~\ref{subsec:objectalgebras}, is that transformation algebras are
frequently used in parsing, as concrete syntax may not always coincide
with its abstract syntax, for instance, the bruijn indices in lambda
calculus, as well as other syntactic sugars. In the Shy
framework~\cite{Zhang2015}, the pattern of transformations is captured
into a template for Java annotation processing, and most of
boilerplate code is automatically generated, hence users only need to
override a few interesting cases, conveniently. It is potentially useful
in our case. By the way, the idea of overriding
existing cases was also proposed in MRM. Our approach allows existing parsing
code to be overridden during inheritance.

>>>>

\paragraph{Parser Combinators} Since Burge~\cite{burge1975} firstly
introduced parser generators, and after Wadler popularized them 
in functional programming~\cite{Wadler1985}, parser combinators have been
more and more popularly developed and used in the research area of
parsing. Many parsing libraries produce recursive descent
parsers by introducing functional monadic
parser combinators~\cite{nott237}. Parsec~\cite{Leijen2001} is 
perhaps the most popular parser combinator library in this line. 
It is widely used in Haskell (with various ``clones'' in other languages)
for context-sensitive grammars with infinite lookahead. Nevertheless,
Parsec users suffer from manual left-recursion elimination, 
high cost for backtracking and longest match composition issues, 
as we discussed in Section~\ref{}. Those limitations make Parsec 
(and similar parsing techniques) inadequate for modular parsing.

@@

\paragraph{Extensibility} Various design patterns~\cite{}, in multiple
languages, have been proposed over the years to address extensibility
problems, such as the Expression Problem~\cite{}. However, to the best
of our knowledge, none of the existing extensibility techniques
address the problem of \emph{modularly 
parsing extensible ASTs}.  There has been
a lot of work on extensible ASTs, in both functional and
object-oriented programming.  For instance, the famous ``Datatypes \`a
la Carte'' (DTC)
paper~\cite{swierstra2008} represents modular ASTs using co-products
of every two functors, which forms a binary tree structure. 
Several variants of DTC have been later proposed~\cite{Bahr2011,Bahr2014,Oliveira2015}
All of that work essentially covers how to traverse and 
consume extensible ASTs. 
Only in Bahr's~\cite{Bahr2011} work a brief mention of \emph{unfolds} 
(which are a class of algorithms that \emph{build} ASTs) are mentioned. 
However Bahr's work does not cover parsing.

In OO languages, there are also many design patterns that achieve 
type-safe extensibility~\cite{torgerson,odersky,Oliveira09,Oliveira12,Wang16}. 
We chose Object Algebras~\cite{Oliveira2012} because the pattern is 
relatively lightweight and makes good use of existing OO features,
such as inheritance, generics and subtyping. As seen throughout the paper, 
the parsing code is concise and expressive using Object Algebras.
In principle our approach is not coupled to the use of Object
Algebras, and variants of our approach could be used with other 
extensibility techniques. Similarly to the work in functional
programming, the extensibility techniques in OOP have focused 
on traversing and processing ASTs. We are unware of any work on OOP
that covers how to do modular parsing for extensible ASTs.

It is worth mentioning that Scala case classes~\cite{} provide a near
solution to the Expression Problem. Case classes are similar to
algebraic datatypes but, unlike algebraic datatypes, new cases can be
modularly added. The reason why case classes are not a complete
solution to the Expression Problem is that they do not enforce
exhaustiveness of pattern matching for extensible operations. In other
words run-time pattern matching errors can happen when writting
extensible code with case classes.  So full static type-safety is not
ensured. Nevertheless case classes are a pragmatic approach, which is
widely used in practice. Therefore modular parsing techniques may be
of value for extensible code using case classes.  The approach we
presented in Section 3~\ref{} can readily be adapted to case classes:
all that the users need to do is to use case classes, instead of
standard OO classes in their code.

\begin{comment}
Moreover, in~\cite{Oliveira2012} the authors have discussed the
composition of algebras. In our parsing approach, a parser consumes an
algebra, which is delegated to return the results, during its process
of parsing. Having a set of algebras, it requires multiple parsing
with several times of invocation, which leads to redundant work.
Instead, algebras are supposed to be composed into one before the
invocation of the parser. Bahr et al. lead a similar discussion
in~\cite{Bahr2011}, where queries (or \textit{catamorphisms}) and
transformations (or \textit{homomorphisms}) are composable. They have
also mentioned the dual process of folds, namely
\textit{anamorphisms}. It is potentially related to our work, as
parsing is a representative kind of unfolds, whereas they only
discussed the composition of a cv-coalgebra and a term homomorphism,
which differs from modular parsing.
\end{comment}

\bruno{There's a problem with the text in 4.2. We need to first 
fix the problem in S4.2 and then come back to the text below and 
adapt it accordingly.}
Another interesting observation, as we have mentioned
in~\ref{subsec:objectalgebras}, is that transformation algebras are
frequently used in parsing, as concrete syntax may not always coincide
with its abstract syntax, for instance, the bruijn indices in lambda
calculus, as well as other syntactic sugars. In the Shy
framework~\cite{Zhang2015}, the pattern of transformations is captured
into a template for Java annotation processing, and most of
boilerplate code is automatically generated, hence users only need to
override a few interesting cases, conveniently. It is potentially
useful in our case. The idea of overriding existing cases was also
proposed in MRM. Our approach allows existing parsing code to be
overridden during inheritance.
